On the probability of finding nonphysical solutions through shadowing\tnoteref{1}

### 2020-12-12 main_els.tex
Intro complete.

We refer to the statistically mismatched shadowing solutions as {\em nonphysical} solutions since they do not represent the behavior of a physically observed solution starting almost everywhere. We identify different ways in which nonphysical solutions occur, adding to the known examples such as periodic orbits embedded within a chaotic attractor. The shadowing lemma \cite{anosov}\cite{bowen} establishes the existence of shadowing orbits corresponding to small perturbations but does not preclude the possibility of nonphysical shadowing orbits (\cite{katok} Theorem 18.1.2).

Owing to the {\em shadowing} property of dynamical systems, it is generally believed that a numerical simulation of chaotic governing equation, which are small perturbations of the true solution, are {\em shadowed} or closely followed by a true solution of the physics.

This also leads us to the conclusion that shadowing-based sensitivities \cite{qiqi-lss} of long-time averages to parameter changes can be incorrect. Thus, the main conjecture of this paper urges us to reconsider the use of chaotic numerical simulations, and statistical sensitivity analysis on these simulations. These implications are particularly important for climate studies, where chaotic models are widely used to make long-term predictions. In particular, three implications of the evidence in this paper may be stated, the third of which harnesses the statistical inconsistency between slightly different models:

In this paper, we show that a large mismatch is possible in the statistical or long-time averages measured in a pair of chaotic solutions to slightly different models.  that  We construct numerical and analytical examples that illustrate this feature of chaotic dynamics. The analysis of the long-term behavior of {\em shadowing} solutions in the majority of the chaotic models constructed in this paper reveals that shadowing solutions are almost surely {\em nonphysical}: they do not represent the behavior of a physically observed solution starting almost everywhere. This also leads us to the conclusion that shadowing-based sensitivities \cite{qiqi-lss} of long-time averages to parameter changes can be incorrect. Thus, the main conjecture of this paper urges us to reconsider the use of chaotic numerical simulations, and statistical sensitivity analysis on these simulations. These implications are particularly important for climate studies, where chaotic models are widely used to make long-term predictions. In particular, three implications of the evidence in this paper may be stated, the third of which harnesses the statistical inconsistency between slightly different models:
>>>>>>> origin/overleaf-2020-12-12-1530
\begin{itemize}
    \item we cannot trust the long-time averages generated by numerical simulations of chaotic processes, even when we eliminate model uncertainties stemming from imperfect modeling of the underlying physics, and statistical noise due to finite-time averaging;
    \item we cannot trust the sensitivities of long-time averages or statistical averages to small parameter changes computed by using shadowing-based methods for sensitivity computation;
	\item it is possible for two models separated by tiny changes in parameter values to have drastically different statistical behavior. This means that the Lorenzian butterfly (a small perturbation) can significantly alter the climate of Texas (statistical/long-term average), and not just its weather (short-term events).
\end{itemize}
The paper is organized as follows. In section \ref{sec:shadowingIntro}, we introduce the dynamical systems concept of shadowing and its applications in the context of chaotic numerical simulations. In section \ref{sec:nonphysical}, we define the notion of physicality and give examples to illustrate the different ways in which to obtain nonphysical solutions. Section \ref{sec:shadowingnonphysical} is dedicated to several examples of perturbed tent maps that have nonphysical shadowing solutions. In the discussion in section \ref{sec:discussion}, we illustrate with an example that a small parameter perturbation can change the statistics significantly. The supplementary material, section \ref{sec:supp}, contains the numerical procedure to generate the probability distributions of the shadowing solutions.

>>>>

We refer to the statistically mismatched shadowing solutions as {\em nonphysical} solutions since they do not represent the behavior of a physically observed solution starting almost everywhere. We identify different ways in which nonphysical solutions occur, adding to the known examples such as periodic orbits embedded within a chaotic attractor. The shadowing lemma \cite{anosov}\cite{bowen} establishes the existence of shadowing orbits corresponding to small perturbations but does not preclude the nonphysicality of these shadowing orbits (\cite{katok} Theorem 18.1.2). One contribution of the present work is the construction of rigorous examples that show that shadowing orbits can be nonphysical, including relating them to previously identified forms of nonphysical orbits, and computing the probability of finding nonphysical shadowing orbits. Another contribution is the interpretation of this finding in the context of chaotic numerical simulations and sensitivity analysis on such simulations.

Owing to the shadowing lemma, it is generally believed that numerical simulations of chaotic governing equations, which are small perturbations of true solutions, are shadowed by true solutions of the governing physics. However, if these shadowing solutions are nonphysical, numerical simulations do not produce the correct long-time averages. Furthermore, the computation of parametric derivatives or {\em linear response} of 
long-time averages by using the idea of shadowing \cite{qiqi-lss}, leads to incorrect values for those derivatives. Thus, the evidence of nonphysicality of shadowing presented in this paper urges us to reconsider the use of chaotic numerical simulations, and statistical sensitivity analysis on these simulations, which are critical for cross-disciplinary applications. These implications are particularly important for climate studies, where chaotic models are widely used to make long-term predictions. In particular, three broad implications of the evidence in this paper may be stated, the third of which harnesses the statistical inconsistency between slightly different models:
\begin{itemize}
    \item we cannot trust the long-time averages generated by numerical simulations of chaotic processes, even when we account for model uncertainties and statistical noise due to finite-time averaging;
    \item we cannot trust the sensitivities of long-time averages or statistical averages to small parameter changes computed by using shadowing-based methods;
	\item it is possible for two models separated by tiny changes in parameter values to have drastically different statistical behavior. This means that the Lorenzian butterfly (a small perturbation) can significantly alter the climate of Texas (statistical/long-term average), and not just its weather (short-term averages).
\end{itemize}
The paper is organized as follows. In section \ref{sec:shadowingIntro}, we introduce the dynamical systems concept of shadowing and its applications in the context of chaotic numerical simulations. In section \ref{sec:nonphysical}, we define the notion of physicality and give examples to illustrate the different ways in which to obtain nonphysical solutions. Section \ref{sec:shadowingnonphysical} is dedicated to several examples of perturbed tent maps that have nonphysical shadowing solutions. In the discussion in section \ref{sec:discussion}, we illustrate with an example that a small parameter perturbation can change the statistics significantly. The supplementary material, section \ref{sec:supp}, contains the numerical procedure to generate the probability distributions of the shadowing solutions.
